---
title: "Metrics Reference: Selection Criteria for K and r"
author: "Yusuke Matsui"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metrics Reference: Selection Criteria for K and r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = FALSE
)
```

## Introduction

This reference document provides detailed explanations of all metrics used for selecting the optimal number of factors (r) and clusters (K) in mixture-based muscle synergy analysis. Each metric has specific strengths and limitations, making it important to understand when and how to use each one.

## Quick Reference Table

| Metric | Purpose | Direction | Best For | Limitations |
|--------|---------|-----------|----------|-------------|
| **CV** | Generalization | Lower | Global r estimation | Computationally expensive |
| **VAF** | Reconstruction | Higher | Minimum r threshold | Cannot detect overfitting |
| **BIC** | Model complexity | Lower | Joint K,r selection | Sensitive to sample size |
| **Smoothness** | Temporal quality | Lower | EMG-specific validation | Domain-specific |
| **VSS** | Interpretability | Higher | Post-rotation evaluation | Requires rotation |
| **PA** | Dimension estimation | Count | Exploratory analysis | Conservative |

## Detailed Metric Descriptions

---

### Cross-Validation (CV)

#### Theoretical Background

Cross-validation estimates the model's ability to generalize to unseen data. In the context of factor analysis, we use leave-one-subject-out CV:

For each subject $i$:
1. Fit model on $N-1$ subjects (excluding $i$)
2. Compute error on held-out subject $i$
3. Average errors across all subjects

#### Mathematical Formulation

**Negative Log-Likelihood CV:**
$$
\text{CV}_{\text{NegLogLik}} = \frac{1}{N} \sum_{i=1}^{N} -\log p(X_i | \hat{\theta}_{-i})
$$

**Frobenius Norm CV:**
$$
\text{CV}_{\text{Frob}} = \frac{1}{N} \sum_{i=1}^{N} \|X_i - \hat{X}_{i,-i}\|_F^2
$$

where $\hat{\theta}_{-i}$ are parameters estimated without subject $i$.

#### Implementation

```{r cv-implementation}
library(synergyMixR)

# CV-based r selection
result_cv <- select_r(
  list_of_data,
  method = "cv",
  rvec = 2:6,
  model_type = "MFA"
)

# View results
print(result_cv)
cat("Selected r:", result_cv$r_selected, "\n")

# Visualize
plot(result_cv$rvec, result_cv$cv_errors, type = "b",
     xlab = "r", ylab = "CV Error",
     main = "Cross-Validation Curve")
abline(v = result_cv$r_selected, col = "red", lty = 2)
```

#### When to Use

- **Primary use case**: Global r estimation before considering cluster structure
- **Data requirements**: Sufficient subjects (N >= 15-20 recommended)
- **Computational note**: Most expensive metric; consider reducing `n_init` for speed

#### Interpretation Guidelines

| CV Curve Pattern | Interpretation | Action |
|------------------|----------------|--------|
| Clear minimum | Strong signal for optimal r | Use minimum |
| Flat curve | r has little effect | Consider smallest r |
| Monotonic decrease | May need larger r range | Expand rvec |
| Noisy curve | Unstable estimation | Increase n_init or subjects |

---

### Variance Accounted For (VAF)

#### Theoretical Background

VAF measures the proportion of total variance in the data explained by the model. It is analogous to $R^2$ in regression.

#### Mathematical Formulation

$$
\text{VAF} = 1 - \frac{\text{SSE}}{\text{SST}} = 1 - \frac{\sum_{i=1}^{N} \|X_i - \hat{X}_i\|_F^2}{\sum_{i=1}^{N} \|X_i - \bar{X}\|_F^2}
$$

where:
- $X_i$ is the observed data for subject $i$
- $\hat{X}_i$ is the reconstructed data
- $\bar{X}$ is the grand mean across all subjects

#### Implementation

```{r vaf-implementation}
# Compute VAF for a fitted model
vaf <- compute_global_vaf_mfa(list_of_data, mfa_fit)
cat("Global VAF:", round(vaf, 4), "\n")

# VAF-based r selection
result_vaf <- select_r_by_vaf_mfa(
  list_of_data,
  K = 3,
  rvec = 2:6,
  vaf_threshold = 0.90  # Select smallest r achieving 90% VAF
)

cat("Selected r:", result_vaf$best_r, "\n")
print(result_vaf$summary)
```

#### VAF Thresholds in Literature

| Domain | Typical Threshold | Reference |
|--------|-------------------|-----------|
| Muscle synergy (locomotion) | 0.90 | Ivanenko et al., 2004 |
| Muscle synergy (reaching) | 0.85-0.90 | d'Avella et al., 2006 |
| General EMG analysis | 0.80-0.85 | Tresch et al., 2006 |

#### Caution: VAF Alone is Insufficient

VAF increases monotonically with r. Choosing r based only on VAF will favor overfitting:

```{r vaf-monotonic}
# Demonstrate VAF's monotonic increase
for (r in 1:6) {
  fit <- mfa_em_fit(list_of_data, K = 3, r = r, n_init = 1)
  vaf_r <- compute_global_vaf_mfa(list_of_data, fit)
  cat("r =", r, ": VAF =", round(vaf_r, 4), "\n")
}
# Output: VAF always increases with r
```

**Solution**: Use VAF as a threshold constraint, not the primary selection criterion.

---

### Bayesian Information Criterion (BIC)

#### Theoretical Background

BIC balances model fit (likelihood) against model complexity (number of parameters). It penalizes complex models to prevent overfitting.

#### Mathematical Formulation

$$
\text{BIC} = -2 \log L(\hat{\theta}) + p \log(n)
$$

where:
- $L(\hat{\theta})$ is the maximized likelihood
- $p$ is the number of free parameters
- $n$ is the sample size (total time points)

For MFA:
$$
p = K \cdot (M \cdot r + 2M) + (K - 1)
$$

- $M \cdot r$: loadings per cluster
- $2M$: means and diagonal noise per cluster
- $K - 1$: mixing proportions

#### Implementation

```{r bic-implementation}
# Compute BIC for a single model
loglik <- mfa_fit$logLik
bic <- compute_BIC_mfa(loglik, K = 3, r = 4, M = 8, N_total_rows = 6000)

# Grid search with BIC selection
grid_result <- select_optimal_K_r_mfa(
  list_of_data,
  Kvec = 1:5,
  rvec = 1:5
)

# View BIC surface
print(grid_result$summary)

# Visualize BIC heatmap
library(ggplot2)
ggplot(grid_result$summary, aes(x = r, y = K, fill = BIC)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1) +
  geom_point(data = grid_result$best_model_info,
             aes(x = r, y = K), color = "red", size = 4)
```

#### BIC vs AIC

| Criterion | Penalty | Tendency |
|-----------|---------|----------|
| BIC | $p \log(n)$ | More parsimonious (prefers simpler models) |
| AIC | $2p$ | Less conservative |

For muscle synergy analysis with moderate sample sizes, BIC is generally preferred.

#### Sample Size Sensitivity

BIC's penalty term $p \log(n)$ grows with sample size. With very large datasets:
- BIC may select very simple models
- Consider AIC or cross-validation as alternatives

---

### Smoothness Penalty

#### Theoretical Background

The smoothness penalty is designed for time-series data like EMG. It measures how "wiggly" the factor scores are over time. When r is too large, models tend to fit high-frequency noise, producing non-smooth factors.

#### Mathematical Formulation

Using second-order differences:
$$
\text{Smoothness} = \sum_{i=1}^{N} \sum_{j=1}^{r} \sum_{t=3}^{T_i} (z_{i,j,t} - 2z_{i,j,t-1} + z_{i,j,t-2})^2
$$

This is equivalent to summing the squared discrete second derivatives of all factor score time series.

#### Implementation

```{r smoothness-implementation}
# Compute factor scores first
mfa_fit <- compute_factor_scores_mfa(mfa_fit, list_of_data)

# Compute smoothness
smoothness <- compute_smoothness_penalty(mfa_fit$factor_scores)
cat("Smoothness penalty:", smoothness, "\n")

# Compare smoothness across r values
for (r in 2:6) {
  fit <- mfa_em_fit(list_of_data, K = 3, r = r, n_init = 1)
  fit <- compute_factor_scores_mfa(fit, list_of_data)
  smooth_r <- compute_smoothness_penalty(fit$factor_scores)
  cat("r =", r, ": Smoothness =", round(smooth_r, 2), "\n")
}
```

#### Interpretation

| Smoothness Level | Interpretation |
|------------------|----------------|
| Low | Smooth, physiologically plausible activations |
| High | Noisy, potentially overfitting high-frequency components |

#### Typical Pattern

As r increases:
1. Initially: smoothness may stay low (capturing true structure)
2. Eventually: smoothness increases sharply (fitting noise)

Look for the "elbow" where smoothness begins to increase rapidly.

#### Muscle Synergy Context

In muscle synergy analysis, factor scores represent muscle activation patterns that should:
- Vary smoothly over the movement cycle
- Not contain high-frequency oscillations (unless physiologically meaningful)
- Reflect coordinated motor control

---

### Very Simple Structure (VSS)

#### Theoretical Background

VSS evaluates the interpretability of factor loadings. The concept of "simple structure" (Thurstone, 1947) suggests that each variable (muscle) should load primarily on one factor (synergy).

#### Mathematical Formulation

**Threshold Method:**
$$
\text{VSS}_{\text{threshold}} = \frac{1}{K \cdot M} \sum_{k=1}^{K} \sum_{m=1}^{M} \mathbb{1}\left[\sum_{j=1}^{r} \mathbb{1}(|\lambda_{k,m,j}| > \tau) = 1\right]
$$

Counts the proportion of muscles with exactly one loading above threshold $\tau$.

**Ratio Method:**
$$
\text{VSS}_{\text{ratio}} = \frac{1}{K \cdot M} \sum_{k=1}^{K} \sum_{m=1}^{M} \frac{\max_j(\lambda_{k,m,j}^2)}{\sum_j \lambda_{k,m,j}^2}
$$

Measures how concentrated each muscle's loading is on its dominant factor.

#### Implementation

```{r vss-implementation}
# Compute VSS
vss_thresh <- compute_vss_criterion(mfa_fit$Lambda, threshold = 0.5, method = "threshold")
vss_ratio <- compute_vss_criterion(mfa_fit$Lambda, method = "ratio")

cat("VSS (threshold):", round(vss_thresh, 4), "\n")
cat("VSS (ratio):", round(vss_ratio, 4), "\n")

# VSS after rotation (recommended)
rotated_fit <- rotate_mfa_model(mfa_fit, rotation = "varimax")
vss_rotated <- compute_vss_criterion(rotated_fit$Lambda, method = "ratio")
cat("VSS after varimax:", round(vss_rotated, 4), "\n")
```

#### Rotation Importance

Unrotated factor loadings often do not exhibit simple structure. Apply varimax (or another rotation) before computing VSS:

```{r vss-rotation-effect}
# Without rotation - often lower VSS
vss_unrot <- compute_vss_criterion(mfa_fit$Lambda, rotate = "none")

# With rotation - typically higher VSS
vss_rot <- compute_vss_criterion(mfa_fit$Lambda, rotate = "varimax")

cat("Unrotated VSS:", round(vss_unrot, 4), "\n")
cat("Rotated VSS:", round(vss_rot, 4), "\n")
```

#### Interpretation Guidelines

| VSS Score | Interpretation |
|-----------|----------------|
| > 0.8 | Excellent simple structure |
| 0.6 - 0.8 | Good simple structure |
| 0.4 - 0.6 | Moderate; some complex loadings |
| < 0.4 | Poor simple structure; difficult to interpret |

---

### Parallel Analysis (PA)

#### Theoretical Background

Parallel analysis compares observed eigenvalues to those expected from random data. Factors are retained only if their eigenvalues exceed the random baseline.

#### Algorithm

1. Compute eigenvalues of pooled covariance matrix: $\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_M$
2. Generate B random permutations of the data
3. For each permutation, compute eigenvalues
4. Calculate 95th percentile of random eigenvalues: $\lambda^{95}_1, ..., \lambda^{95}_M$
5. Select r = max{j : $\lambda_j > \lambda^{95}_j$}

#### Implementation

```{r pa-implementation}
# Global MFA-based r estimation
result_pa <- select_r_global_mfa(
  list_of_data,
  rvec = 1:8
)

cat("Selected r:", result_pa$r_global, "\n")

# View results
print(result_pa)
```

#### Characteristics

**Strengths:**
- Objective, data-driven threshold
- No arbitrary VAF cutoff needed
- Good for exploratory analysis

**Limitations:**
- Conservative (may underestimate r)
- Ignores cluster structure
- Assumes single population

---

## Metric Comparison and Selection Strategy

### Single Metric Usage

| Metric | When to Use Alone |
|--------|-------------------|
| CV | Global r estimation with sufficient data |
| BIC | Joint K,r selection via grid search |
| PA | Quick exploratory analysis |

### Recommended Combinations

#### Conservative Selection (Avoid Overfitting)

```{r conservative}
# Prioritize BIC and smoothness
result <- select_r_multicriteria_mfa(
  list_of_data, K = 3, rvec = 2:6,
  weights = c(BIC = 2, VAF = 1, smoothness = 2, VSS = 1)
)
```

#### Interpretability Focus

```{r interpretability}
# Prioritize VSS
result <- select_r_multicriteria_mfa(
  list_of_data, K = 3, rvec = 2:6,
  weights = c(BIC = 1, VAF = 1, smoothness = 1, VSS = 2)
)
```

#### Balanced Approach (Recommended Default)

```{r balanced}
result <- select_r_multicriteria_mfa(
  list_of_data, K = 3, rvec = 2:6,
  vaf_threshold = 0.85,
  weights = c(BIC = 1, VAF = 1, smoothness = 1, VSS = 1)
)
```

### Decision Framework

```
┌─────────────────────────────────────────────────────────────┐
│ Step 1: Establish baseline with global CV                   │
│ r_cv = select_r(method = "cv", ...)$r_selected               │
└────────────────────┬────────────────────────────────────────┘
                     │
                     ▼
┌─────────────────────────────────────────────────────────────┐
│ Step 2: Check VAF threshold                                  │
│ If VAF(r_cv) < 0.85 → consider larger r                     │
│ If VAF(r_cv) > 0.95 → may be overfitting, check smoothness  │
└────────────────────┬────────────────────────────────────────┘
                     │
                     ▼
┌─────────────────────────────────────────────────────────────┐
│ Step 3: Validate with secondary metrics                      │
│ - Check smoothness (should not spike)                       │
│ - Check VSS after rotation (should be reasonable)           │
└────────────────────┬────────────────────────────────────────┘
                     │
                     ▼
┌─────────────────────────────────────────────────────────────┐
│ Step 4: Joint K,r selection with BIC                        │
│ Use focused rvec around validated r                         │
└─────────────────────────────────────────────────────────────┘
```

## Summary

### Key Takeaways

1. **No single metric is perfect** - each has strengths and limitations
2. **CV is the gold standard** for generalization but is computationally expensive
3. **VAF should be a constraint**, not the primary criterion
4. **BIC works well for joint K,r selection** with moderate sample sizes
5. **Smoothness is EMG-specific** - leverage temporal structure of your data
6. **VSS requires rotation** - always apply varimax before evaluation
7. **PA is conservative** - useful for exploratory analysis

### Recommended Workflow

1. Start with global CV or PA for r estimation
2. Set VAF >= 0.85 as minimum threshold
3. Use multi-criteria selection with balanced weights
4. Validate final selection with smoothness and VSS
5. Visualize loadings and scores for interpretability

## References

- Ivanenko, Y. P., et al. (2004). Five basic muscle activation patterns account for muscle activity during human locomotion. *Journal of Physiology*.
- d'Avella, A., et al. (2006). Control of fast-reaching movements by muscle synergy combinations. *Journal of Neuroscience*.
- Tresch, M. C., et al. (2006). Matrix factorization algorithms for the identification of muscle synergies. *Journal of Neurophysiology*.
- Revelle, W., & Rocklin, T. (1979). Very simple structure: An alternative procedure for estimating the optimal number of interpretable factors. *Multivariate Behavioral Research*.
- Horn, J. L. (1965). A rationale and test for the number of factors in factor analysis. *Psychometrika*.

## Session Info

```{r session}
sessionInfo()
```
