---
title: "Comprehensive Guide to K and r Model Selection"
author: "Yusuke Matsui"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive Guide to K and r Model Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
 fig.height = 6,
  eval = FALSE
)
```

## Introduction

This vignette provides a comprehensive guide to selecting the optimal number of clusters (K) and factors (r) in mixture-based muscle synergy analysis. While individual selection methods are documented elsewhere, this guide focuses on:

1. **Two-stage K,r selection**: A stable approach for joint optimization
2. **Post-hoc evaluation**: Tools for comparing multiple model candidates
3. **Integrated workflow**: End-to-end decision-making process for analysts

## Load Package and Example Data

```{r load}
library(synergyMixR)

# Generate synthetic data
sim_data <- generate_synergy_data(
  N = 60, K = 3, r = 4, M = 8,
  T_each = 100,
  cluster_sep_spatial = 1.0,
  seed = 42
)

list_of_data <- sim_data$list_of_data
z_true <- sim_data$true_cluster
cat("True: K =", 3, ", r =", 4, "\n")
```

## Part 1: Two-Stage K,r Selection

### Why Two-Stage?

In mixture models, K (number of clusters) and r (number of factors) have strong interactions. Simultaneously searching over all (K, r) combinations can be:

- **Computationally expensive**: O(|Kvec| × |rvec|) model fits
- **Unstable**: Different initializations may prefer different (K, r) pairs
- **Confounded**: Increasing K can compensate for inadequate r, and vice versa

The two-stage approach addresses these issues by:

1. First fixing r and optimizing K
2. Then fixing K at the optimal value and optimizing r

### Algorithm

```
Stage 1: K Selection (r fixed)
├── Fix r = r_fixed (default: 2)
├── For each K in Kvec:
│   └── Fit model, compute BIC
└── Select best_K = argmin(BIC)

Stage 2: r Selection (K fixed)
├── Fix K = best_K
├── For each r in rvec:
│   └── Fit model, compute BIC
└── Select best_r = argmin(BIC)
```

### Usage Example

```{r twostage-mfa}
# Two-stage selection for MFA
result_2stage <- select_optimal_K_r_mfa_twostage(
  list_of_data,
  Kvec = 2:5,      # Candidates for K
  rvec = 1:6,      # Candidates for r
  r_fixed = 2,     # Fixed r for Stage 1
  max_iter = 50,
  n_init = 3,
  use_kmeans_init = TRUE
)

# Results
cat("=== Two-Stage Selection Results ===\n")
cat("Stage 1 (K search with r =", 2, "):\n")
print(result_2stage$stage1_summary)

cat("\nStage 2 (r search with K =", result_2stage$best_K, "):\n")
print(result_2stage$stage2_summary)

cat("\nFinal selection: K =", result_2stage$best_K, ", r =", result_2stage$best_r, "\n")
```

### For MPCA Models

```{r twostage-mpca}
# Two-stage selection for MPCA
result_2stage_mpca <- select_optimal_K_r_mpca_twostage(
  list_of_data,
  Kvec = 2:5,
  rvec = 1:6,
  r_fixed = 2,
  method = "EM"
)

cat("MPCA: K =", result_2stage_mpca$best_K, ", r =", result_2stage_mpca$best_r, "\n")
```

### Choosing r_fixed

The `r_fixed` parameter in Stage 1 affects the K selection. Guidelines:

| r_fixed | When to Use |
|---------|-------------|
| 2 (default) | Safe default for most applications |
| 3-4 | When you expect moderate synergy complexity |
| 1 | When data is very noisy or sample size is small |

If you have prior knowledge about r (e.g., from literature or global r estimation), use that value:

```{r rfixed-from-global}
# Use global r estimate as r_fixed
global_result <- select_r_global_mfa(list_of_data, rvec = 2:6)
r_estimated <- global_result$r_global

result_informed <- select_optimal_K_r_mfa_twostage(
  list_of_data,
  Kvec = 2:5,
  rvec = 1:6,
  r_fixed = r_estimated  # Use estimated r
)
```

### Comparison: Two-Stage vs Grid Search

| Method | Complexity | Stability | Optimality |
|--------|------------|-----------|------------|
| Grid Search | O(|Kvec| × |rvec|) | Low (many local minima) | High (explores all) |
| Two-Stage | O(|Kvec| + |rvec|) | High (sequential) | Moderate (depends on r_fixed) |

**Recommendation**:
- Use **two-stage** for initial exploration or when computation is limited
- Use **grid search** for final model selection when resources allow

## Part 2: Post-Hoc Evaluation

After running a grid search, you may want to compare multiple (K, r) candidates beyond just BIC. The post-hoc evaluation functions provide additional metrics.

### posthoc_mfa_evaluation()

This function computes additional metrics for all models from a grid search:

```{r posthoc}
# First, run grid search
grid_result <- select_optimal_K_r_mfa(
  list_of_data,
  Kvec = 2:4,
  rvec = 2:5,
  n_init = 3
)

# Post-hoc evaluation
df_posthoc <- posthoc_mfa_evaluation(list_of_data, grid_result)

# View results (sorted by BIC)
print(df_posthoc)
```

### Output Columns

| Column | Description | Good Value |
|--------|-------------|------------|
| `K` | Number of clusters | - |
| `r` | Number of factors | - |
| `logLik` | Log-likelihood | Higher is better |
| `BIC` | Bayesian Information Criterion | Lower is better |
| `GlobalVAF` | Variance Accounted For | > 0.85 |
| `minClusterSize` | Smallest cluster size | > N/(2K) |
| `maxClusterSize` | Largest cluster size | < N - N/(2K) |

### Interpreting Cluster Sizes

Extreme cluster size imbalance can indicate:

- **Overfitting**: One cluster captures outliers
- **Wrong K**: True K may be different
- **Poor initialization**: Try more `n_init`

```{r cluster-balance}
# Check for balanced clusters
df_balanced <- df_posthoc[
  df_posthoc$minClusterSize >= 10 &  # At least 10 subjects per cluster
  df_posthoc$GlobalVAF >= 0.85,      # Adequate reconstruction
]

cat("Candidates with balanced clusters and good VAF:\n")
print(df_balanced)
```

### get_model_by_K_r()

Retrieve a specific model for detailed analysis:

```{r get-model}
# Get a model that's not BIC-optimal but has other desirable properties
# e.g., higher VAF, more balanced clusters
alt_model <- get_model_by_K_r(grid_result, K_target = 3, r_target = 4)

if (!is.null(alt_model)) {
  # Analyze this model
  alt_model <- compute_factor_scores_mfa(alt_model, list_of_data)

  # Visualize synergy patterns
  plot_cluster_synergy_loadings_mfa(alt_model, plot_type = "heatmap")

  # Check factor score smoothness
  smoothness <- compute_smoothness_penalty(alt_model$factor_scores)
  cat("Smoothness penalty:", smoothness, "\n")
}
```

### Comparing Multiple Candidates

```{r compare-candidates}
# Compare top 3 candidates by different criteria
top_by_bic <- df_posthoc[order(df_posthoc$BIC), ][1:3, ]
top_by_vaf <- df_posthoc[order(-df_posthoc$GlobalVAF), ][1:3, ]

cat("Top 3 by BIC:\n")
print(top_by_bic[, c("K", "r", "BIC", "GlobalVAF")])

cat("\nTop 3 by VAF:\n")
print(top_by_vaf[, c("K", "r", "BIC", "GlobalVAF")])

# If different, investigate both
```

## Part 3: Integrated Workflow

This section presents the recommended end-to-end workflow for analysts.

### Workflow Diagram

```
┌─────────────────────────────────────────────────────────────────┐
│ Phase 1: Preliminary Investigation                              │
├─────────────────────────────────────────────────────────────────┤
│ 1. Global r estimation (K-independent)                         │
│    select_r_global_mfa(list_of_data, rvec = 2:6)               │
│    → Obtain r_global as baseline                               │
│                                                                 │
│ 2. Visualize CV curve                                          │
│    plot(result, type = "global_cv")                            │
└────────────────────┬────────────────────────────────────────────┘
                     │
                     ▼
┌─────────────────────────────────────────────────────────────────┐
│ Phase 2: Joint K,r Selection                                    │
├─────────────────────────────────────────────────────────────────┤
│ Option A: Grid Search (comprehensive)                           │
│   select_optimal_K_r_mfa(Kvec = 1:5, rvec = r_global ± 1)      │
│                                                                 │
│ Option B: Two-Stage (efficient)                                 │
│   select_optimal_K_r_mfa_twostage(Kvec = 2:5, rvec = 1:6,      │
│                                    r_fixed = r_global)         │
│                                                                 │
│ Option C: Hybrid r-Selection (for fixed K)                      │
│   select_r_mixture(K = 3, rvec = 2:6, mode = "hybrid")         │
└────────────────────┬────────────────────────────────────────────┘
                     │
                     ▼
┌─────────────────────────────────────────────────────────────────┐
│ Phase 3: Diagnostics and Visualization                          │
├─────────────────────────────────────────────────────────────────┤
│ 1. Summary statistics                                           │
│    print(result); summary(result)                               │
│                                                                 │
│ 2. Visualizations                                               │
│    plot(result, type = "overview")                              │
│    plot_cluster_synergy_loadings_mfa(best_model)               │
│    plot_all_factor_scores_mfa(best_model, list_of_data)        │
│                                                                 │
│ 3. Check for issues                                             │
│    - Boundary r values                                          │
│    - Unbalanced clusters                                        │
│    - Non-smooth factor scores                                   │
└────────────────────┬────────────────────────────────────────────┘
                     │
                     ▼
┌─────────────────────────────────────────────────────────────────┐
│ Phase 4: Post-Hoc Evaluation and Final Decision                 │
├─────────────────────────────────────────────────────────────────┤
│ 1. Compare multiple candidates                                   │
│    df <- posthoc_mfa_evaluation(list_of_data, grid_result)     │
│                                                                 │
│ 2. Retrieve alternative models                                   │
│    alt <- get_model_by_K_r(grid_result, K = 3, r = 4)          │
│                                                                 │
│ 3. Final decision based on:                                     │
│    - BIC vs VAF trade-off                                       │
│    - Cluster balance                                            │
│    - Physiological interpretability                             │
│    - Factor score smoothness                                    │
└─────────────────────────────────────────────────────────────────┘
```

### Complete Example

```{r complete-workflow}
# ========================================
# Phase 1: Preliminary Investigation
# ========================================

cat("=== Phase 1: Global r Estimation ===\n")
global_result <- select_r_global_mfa(
  list_of_data,
  rvec = 2:6,
  criterion = "CV"
)
r_global <- global_result$r_global
cat("Estimated r_global:", r_global, "\n")

# ========================================
# Phase 2: Joint K,r Selection
# ========================================

cat("\n=== Phase 2: Grid Search ===\n")
# Focus search around r_global
rvec_focused <- max(1, r_global - 1):min(8, r_global + 2)

grid_result <- select_optimal_K_r_mfa(
  list_of_data,
  Kvec = 1:5,
  rvec = rvec_focused,
  n_init = 5,
  use_kmeans_init = TRUE
)

best_K <- grid_result$best_model_info$K
best_r <- grid_result$best_model_info$r
cat("BIC-optimal: K =", best_K, ", r =", best_r, "\n")

# ========================================
# Phase 3: Diagnostics
# ========================================

cat("\n=== Phase 3: Diagnostics ===\n")
best_model <- grid_result$best_model_info$model

# Compute factor scores
best_model <- compute_factor_scores_mfa(best_model, list_of_data)

# Check cluster sizes
cluster_sizes <- table(best_model$z)
cat("Cluster sizes:", paste(cluster_sizes, collapse = ", "), "\n")

# Check VAF
vaf <- compute_global_vaf_mfa(list_of_data, best_model)
cat("Global VAF:", round(vaf, 4), "\n")

# Check smoothness
smoothness <- compute_smoothness_penalty(best_model$factor_scores)
cat("Smoothness penalty:", round(smoothness, 2), "\n")

# Visualize
plot_cluster_synergy_loadings_mfa(best_model, plot_type = "heatmap")

# ========================================
# Phase 4: Post-Hoc Evaluation
# ========================================

cat("\n=== Phase 4: Post-Hoc Comparison ===\n")
df_posthoc <- posthoc_mfa_evaluation(list_of_data, grid_result)
print(head(df_posthoc, 10))

# Final decision checklist
cat("\n=== Final Decision Checklist ===\n")
cat("[ ] VAF >= 0.85:", vaf >= 0.85, "\n")
cat("[ ] BIC is optimal or near-optimal\n")
cat("[ ] Clusters are reasonably balanced:",
    min(cluster_sizes) >= length(list_of_data) / (2 * best_K), "\n")
cat("[ ] r is not at boundary:",
    best_r > min(rvec_focused) & best_r < max(rvec_focused), "\n")
cat("[ ] Synergy loadings are interpretable: (visual inspection)\n")
cat("[ ] Factor scores are smooth: (visual inspection)\n")
```

### Decision Checklist

Use this checklist for final model selection:

| Check | How to Verify | Good Status |
|-------|---------------|-------------|
| VAF sufficient | `compute_global_vaf_mfa()` | >= 0.85-0.90 |
| BIC near optimal | `grid_result$summary` | Within 10 of minimum |
| Clusters balanced | `table(model$z)` | min >= N/(2K) |
| r not at boundary | Check `best_r` vs `rvec` | Not at edges |
| Loadings interpretable | `plot_cluster_synergy_loadings_mfa()` | Clear structure |
| Scores smooth | `plot_all_factor_scores_mfa()` | No high-frequency noise |

### When to Reconsider

**Scenario 1: r at boundary**
```{r reconsider-boundary}
if (best_r == max(rvec_focused)) {
  cat("Warning: r at upper boundary. Expand rvec.\n")
  # Re-run with expanded range
}
```

**Scenario 2: Unbalanced clusters**
```{r reconsider-balance}
if (min(cluster_sizes) < 5) {
  cat("Warning: Very small cluster. Consider:\n")
  cat("  - Reducing K\n")
  cat("  - Checking for outliers\n")
  cat("  - Using more initializations\n")
}
```

**Scenario 3: Low VAF**
```{r reconsider-vaf}
if (vaf < 0.80) {
  cat("Warning: Low VAF. Consider:\n")
  cat("  - Increasing r\n")
  cat("  - Checking data quality\n")
  cat("  - Reducing noise in preprocessing\n")
}
```

## Summary

This guide covered three key aspects of model selection:

1. **Two-Stage Selection**: Efficient, stable approach for joint K,r optimization
   - Use `select_optimal_K_r_mfa_twostage()`
   - Set `r_fixed` based on global r estimate or domain knowledge

2. **Post-Hoc Evaluation**: Compare candidates beyond BIC
   - Use `posthoc_mfa_evaluation()` for comprehensive metrics
   - Use `get_model_by_K_r()` to retrieve specific models

3. **Integrated Workflow**: Four-phase decision process
   - Phase 1: Preliminary investigation (global r)
   - Phase 2: Joint K,r selection (grid or two-stage)
   - Phase 3: Diagnostics and visualization
   - Phase 4: Post-hoc comparison and final decision

**Key Recommendations**:

- Start with global r estimation to inform search range
- Use focused grid search around estimated r
- Always check cluster balance and VAF
- Visualize synergy patterns for interpretability
- Document your decision rationale

## See Also

- `vignette("r_selection_framework")`: Detailed r-selection methods
- `vignette("metrics_reference")`: Understanding each metric
- `vignette("tutorial")`: Basic MFA/MPCA usage

## Session Info

```{r session}
sessionInfo()
```
