---
title: "Quickstart: Mixture-Based Muscle Synergy Analysis"
author: "Yusuke Matsui"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart: Mixture-Based Muscle Synergy Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This quickstart guide demonstrates the basic workflow for mixture-based muscle synergy analysis using the `synergyMixR` package. We will show a minimal example that:

1. Generates synthetic EMG data using `generate_synergy_data()`
2. Fits a Mixture Factor Analysis (MFA) model using `mfa_em_fit()`
3. Visualizes the synergy loadings using `plot_cluster_synergy_loadings_mfa()`

## Load the Package

```{r load-package}
library(synergyMixR)
library(ggplot2)  # Required for visualization functions
library(tidyr)    # Required for plot_all_factor_scores_mfa
```

## Step 1: Generate Synthetic Data

We start by generating synthetic EMG data with known cluster structure. This simulates a heterogeneous population where different subjects exhibit distinct muscle synergy patterns.

```{r simulate-data}
# Generate synthetic data
# N = 30 subjects, K = 2 clusters, r = 3 synergies, M = 8 muscles
sim_data <- generate_synergy_data(
  N = 30,           # Number of subjects
  K = 2,            # Number of clusters (subgroups)
  r = 3,            # Number of synergies per cluster
  M = 8,            # Number of muscles (EMG channels)
  T_each = 100,     # Time points per subject
  cluster_sep_spatial = 1.0,    # Spatial structure separation
  cluster_sep_temporal = 1.0,   # Temporal coordination separation
  cluster_sep_stability = 1.0,  # Stability-adaptability separation
  seed = 123        # For reproducibility
)

# Extract the data and true cluster assignments
list_of_data <- sim_data$list_of_data
z_true <- sim_data$true_cluster

# Check the structure
cat("Number of subjects:", length(list_of_data), "\n")
cat("Data dimensions for subject 1:", dim(list_of_data[[1]]), "\n")
cat("True cluster assignments:", table(z_true), "\n")
```

## Step 2: Fit MFA Model

Now we fit a Mixture Factor Analysis model to identify the underlying synergy patterns and cluster structure.

```{r fit-mfa}
# Fit MFA model with K=2 clusters and r=3 synergies
fit_mfa <- mfa_em_fit(
  list_of_data = list_of_data,
  K = 2,              # Number of clusters to fit
  r = 3,              # Number of synergies per cluster
  max_iter = 50,      # Maximum EM iterations
  nIterFA = 5,        # Sub-iterations for factor updates
  tol = 1e-3,         # Convergence tolerance
  n_init = 3,         # Number of random initializations
  use_kmeans_init = TRUE  # Use k-means for initialization
)

# Check the results
cat("Converged log-likelihood:", fit_mfa$logLik, "\n")
cat("Estimated cluster assignments:", table(fit_mfa$z), "\n")

# Compare with true assignments (using Adjusted Rand Index)
ari <- mclust::adjustedRandIndex(fit_mfa$z, z_true)
cat("Adjusted Rand Index (clustering accuracy):", round(ari, 3), "\n")
```

## Step 3: Visualize Synergy Loadings

Finally, we visualize the identified muscle synergy patterns for each cluster.

```{r plot-loadings, fig.width=8, fig.height=6}
# Plot synergy loadings as a heatmap
plot_cluster_synergy_loadings_mfa(fit_mfa, plot_type = "heatmap")
```

The heatmap shows the synergy loadings (Lambda matrices) for each cluster. Each row represents a muscle, and each column represents a synergy. The color intensity indicates the strength of each muscle's contribution to each synergy.

## Additional Visualizations

You can also visualize the factor scores (time-varying synergy activations):

```{r plot-scores, fig.width=8, fig.height=6}
# Compute factor scores
fit_mfa <- compute_factor_scores_mfa(fit_mfa, list_of_data)

# Plot factor scores for all subjects
plot_all_factor_scores_mfa(fit_mfa, list_of_data, overlay_subjects = TRUE)
```

## Next Steps

For more advanced usage, including:

- Automatic selection of optimal K and r using BIC
- Post-hoc rotation of synergy patterns
- Comparison between MFA and MPCA methods
- Detailed simulation studies

Please see the main tutorial vignette:

```{r eval=FALSE}
vignette("tutorial", package = "synergyMixR")
```

## Summary

This quickstart demonstrated the basic three-step workflow:

1. **Generate or load data**: Use `generate_synergy_data()` for synthetic data or prepare your own EMG data as a list of matrices
2. **Fit the model**: Use `mfa_em_fit()` to identify clusters and synergy patterns
3. **Visualize results**: Use `plot_cluster_synergy_loadings_mfa()` and related functions to interpret the results

The mixture-based approach is particularly useful when you suspect that your population contains distinct subgroups with different synergy patterns, rather than assuming all subjects share the same underlying structure.
