---
title: "Unified r-Selection Framework for Mixture Models"
author: "Yusuke Matsui"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unified r-Selection Framework for Mixture Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = FALSE  # Set to TRUE when package is fully built
)
```

## Introduction

Selecting the optimal number of factors (r) is a critical step in mixture-based muscle synergy analysis. The `synergyMixR` package provides a unified framework for r-selection that integrates three complementary approaches:

1. **Global-first**: Estimate r from pooled data before considering cluster structure
2. **Clusterwise**: Evaluate r separately for each cluster, then aggregate
3. **Hybrid**: Combine global and clusterwise approaches for robust selection

This vignette demonstrates how to use the unified `select_r_mixture()` function and interpret the results using the built-in visualization methods.

## Load the Package

```{r load-package}
library(synergyMixR)
```

## Generate Example Data

We'll use synthetic data to demonstrate the r-selection framework. This allows us to know the true r value and evaluate how well each method recovers it.

```{r simulate-data}
# Generate synthetic data with known structure
# N = 50 subjects, K = 3 clusters, r = 4 true synergies, M = 8 muscles
sim_data <- generate_synergy_data(
  N = 50,
  K = 3,
  r = 4,  # True number of synergies
  M = 8,
  T_each = 100,
  cluster_sep_spatial = 1.0,
  seed = 123
)

list_of_data <- sim_data$list_of_data
z_true <- sim_data$true_cluster

cat("Generated data with", length(list_of_data), "subjects\n")
cat("True cluster sizes:", table(z_true), "\n")
cat("True r = 4\n")
```

## Mode 1: Global-First Selection

The global-first mode estimates r by pooling all data together, ignoring cluster structure. This provides a fast, cluster-independent baseline estimate.

### When to Use Global-First Mode

- When you want a quick, stable estimate
- As a starting point before more detailed analysis
- When cluster structure is weak or uncertain
- For computational efficiency with large datasets

### Usage Example

```{r global-first}
# Estimate r using global-first mode
result_global <- select_r_mixture(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,  # Candidate r values to evaluate
  mode = "global-first",
  model_type = "MFA",
  global_model = "PPCA",  # Use PPCA for global estimation
  global_criterion = "CV",  # Cross-validation
  global_folds = 5,
  verbose = TRUE
)

# View results
print(result_global)
summary(result_global)
```

### Visualize Results

```{r plot-global, fig.width=7, fig.height=5}
# Plot CV curve
plot(result_global, type = "global_cv")

# Or use overview (same for global-first mode)
plot(result_global)
```

The CV curve shows how cross-validation error varies with r. The selected r_global minimizes this error.

## Mode 2: Clusterwise Selection

The clusterwise mode evaluates r separately for each cluster, then aggregates the results. This allows different clusters to have different optimal r values.

### When to Use Clusterwise Mode

- When you suspect different clusters may require different numbers of synergies
- When cluster structure is well-defined
- When you want cluster-specific insights
- For detailed analysis of heterogeneous populations

### Usage Example

```{r clusterwise}
# Estimate r using clusterwise mode
result_cluster <- select_r_mixture(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  mode = "clusterwise",
  model_type = "MFA",
  criteria = c("VAF", "VSS", "smoothness"),  # Multiple criteria
  agg_method = "max",  # Take maximum r across clusters
  weights = c(VAF = 0.4, VSS = 0.4, smoothness = 0.2),
  verbose = TRUE
)

# View results
print(result_cluster)
summary(result_cluster)

# Check cluster-specific r values
cat("Cluster-specific r values:\n")
print(result_cluster$r_cluster)
cat("Aggregated r_final:", result_cluster$r_final, "\n")
```

### Aggregation Methods

The `agg_method` parameter controls how cluster-specific r values are combined:

- **"max"** (default): Conservative approach, takes the maximum r across clusters
- **"mean"**: Average r across clusters (rounded to nearest integer)
- **"min"**: Minimum r across clusters (most parsimonious)
- **"majority"**: Most frequently selected r value

```{r agg-methods}
# Compare different aggregation methods
methods <- c("max", "mean", "min", "majority")
results <- lapply(methods, function(method) {
  res <- select_r_mixture(
    list_of_data = list_of_data,
    K = 3,
    rvec = 2:6,
    mode = "clusterwise",
    model_type = "MFA",
    agg_method = method,
    verbose = FALSE
  )
  c(method = method, r_final = res$r_final)
})

cat("Comparison of aggregation methods:\n")
do.call(rbind, results)
```

### Visualize Clusterwise Results

```{r plot-clusterwise, fig.width=10, fig.height=8}
# Overview: shows cluster metrics heatmap and cluster scores
plot(result_cluster)

# Individual plot types
plot(result_cluster, type = "cluster_metrics")  # Heatmap
plot(result_cluster, type = "cluster_scores")   # Line plots
```

The heatmap shows how each cluster's composite score varies with r. Blue squares mark the selected r for each cluster.

## Mode 3: Hybrid Selection (Recommended)

The hybrid mode combines the best of both approaches: it uses global estimation to get a stable initialization, then refines with cluster-specific evaluation.

### When to Use Hybrid Mode

- **Recommended as default** for most applications
- Combines stability of global estimation with flexibility of clusterwise evaluation
- Uses r_global as initialization for clusterwise fitting (Phase 3 feature)
- Provides comprehensive diagnostic information

### Usage Example

```{r hybrid}
# Estimate r using hybrid mode
result_hybrid <- select_r_mixture(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  mode = "hybrid",
  model_type = "MFA",
  global_model = "PPCA",
  global_criterion = "CV",
  criteria = c("VAF", "VSS", "smoothness"),
  agg_method = "max",  # Can also use "global" to skip clusterwise
  verbose = TRUE
)

# View results
print(result_hybrid)
summary(result_hybrid)

# Compare global vs cluster-specific estimates
cat("\nComparison:\n")
cat("Global r:", result_hybrid$r_global, "\n")
cat("Cluster-specific r:", result_hybrid$r_cluster, "\n")
cat("Final r:", result_hybrid$r_final, "\n")
```

### Special Aggregation: "global"

In hybrid mode, you can use `agg_method = "global"` to directly use r_global without clusterwise refinement:

```{r hybrid-global}
result_hybrid_simple <- select_r_mixture(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  mode = "hybrid",
  model_type = "MFA",
  agg_method = "global",  # Use r_global directly
  verbose = TRUE
)

cat("Using r_global directly:", result_hybrid_simple$r_final, "\n")
```

### Visualize Hybrid Results

```{r plot-hybrid, fig.width=10, fig.height=8}
# Overview: shows all available plots
plot(result_hybrid)

# Individual plot types
plot(result_hybrid, type = "global_cv")        # CV curve
plot(result_hybrid, type = "r_comparison")     # Bar plot comparing r values
plot(result_hybrid, type = "cluster_metrics")  # Heatmap
plot(result_hybrid, type = "cluster_scores")   # Line plots
```

The r_comparison plot is particularly useful in hybrid mode, showing how r_global compares to cluster-specific estimates and the final r_final.

## Comparison: MFA vs MPCA

The framework supports both Mixture Factor Analysis (MFA) and Mixture PCA (MPCA). Here's how to use MPCA:

```{r mpca}
# Use MPCA instead of MFA
result_mpca <- select_r_mixture(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  mode = "hybrid",
  model_type = "MPCA",  # Change to MPCA
  global_model = "PPCA",  # PPCA for global estimation
  verbose = TRUE
)

print(result_mpca)
plot(result_mpca)
```

**When to use MFA vs MPCA:**

- **MFA**: More flexible, allows for correlated noise (Psi diagonal)
- **MPCA**: Simpler, assumes isotropic noise (single sigma^2)
- **Recommendation**: Start with MFA for muscle synergy analysis

## Advanced: Direct Use of Clusterwise Functions

For more control, you can directly use the underlying clusterwise functions:

```{r direct-clusterwise}
# Direct call to clusterwise function
result_direct <- select_r_clusterwise_mfa(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  mode = "clusterwise",
  criteria = c("VAF", "VSS"),
  agg_method = "max",
  r_init = 4,  # Specify initial r for clustering (Phase 3 feature)
  verbose = TRUE
)

cat("Selected r:", result_direct$agg_r, "\n")
```

The `r_init` parameter (added in Phase 3) allows you to specify the initial r value used for the first-stage clustering. This is automatically set to r_global in hybrid mode.

## Best Practices and Recommendations

### 1. Start with Hybrid Mode

```{r best-practice-1}
# Recommended default workflow
result <- select_r_mixture(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  mode = "hybrid",
  model_type = "MFA",
  verbose = TRUE
)
```

### 2. Use Appropriate rvec Range

- Start with a reasonable range (e.g., 2:6 for muscle synergies)
- Ensure the range includes the expected true r
- Avoid very large ranges to reduce computation time

### 3. Interpret Multiple Criteria

```{r best-practice-3}
# Check cluster-specific metrics
if (!is.null(result$diagnostics$cluster_metrics)) {
  for (k in 1:3) {
    cat("\nCluster", k, "metrics:\n")
    print(result$diagnostics$cluster_metrics[[k]])
  }
}
```

### 4. Validate with Visualization

Always visualize the results to ensure the selection makes sense:

```{r best-practice-4}
plot(result)  # Overview
plot(result, type = "r_comparison")  # Compare estimates
```

### 5. Consider Computational Cost

- **Global-first**: Fastest, O(N)
- **Clusterwise**: Moderate, O(K × |rvec|)
- **Hybrid**: Slowest, O(N + K × |rvec|), but most comprehensive

For large datasets, consider:
- Using `mc_cores > 1` for parallelization
- Reducing `n_init` (number of random initializations)
- Starting with global-first for quick exploration

## Troubleshooting

### Issue: Different modes give very different r values

**Solution**: This suggests cluster structure is important. Use hybrid mode and examine cluster-specific estimates:

```{r troubleshoot-1}
# Check if clusters have different optimal r
cat("Cluster-specific r values:", result_hybrid$r_cluster, "\n")
cat("Global r:", result_hybrid$r_global, "\n")

# Visualize the differences
plot(result_hybrid, type = "r_comparison")
```

### Issue: Selected r is at the boundary of rvec

**Solution**: Expand your rvec range:

```{r troubleshoot-2}
# If r_final = max(rvec), try larger range
if (result$r_final == max(rvec)) {
  result_expanded <- select_r_mixture(
    list_of_data = list_of_data,
    K = 3,
    rvec = 2:8,  # Expanded range
    mode = "hybrid",
    verbose = TRUE
  )
}
```

### Issue: Clusterwise mode gives NA or warnings

**Solution**: Check for degenerate clusters (too few subjects or low VAF):

```{r troubleshoot-3}
# Run with verbose=TRUE to see warnings
result_verbose <- select_r_mixture(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  mode = "clusterwise",
  verbose = TRUE  # Shows degeneracy warnings
)

# Check cluster sizes
if (!is.null(result_verbose$initial_fit)) {
  cat("Cluster sizes:", table(result_verbose$initial_fit$z), "\n")
}
```

## Multi-Criteria Selection

When you want to combine multiple metrics with custom weights, use `select_r_multicriteria_mfa()`:

```{r multicriteria}
# Multi-criteria selection with custom weights
result_multi <- select_r_multicriteria_mfa(
  list_of_data = list_of_data,
  K = 3,
  rvec = 2:6,
  vaf_threshold = 0.85,  # Minimum VAF required
  weights = c(BIC = 1, VAF = 1, smoothness = 1, VSS = 1)
)

# View detailed summary
print(result_multi$summary)
cat("Selected r:", result_multi$best_r, "\n")
```

### Understanding the Composite Score

The multi-criteria approach combines four metrics:

1. **BIC** (lower is better): Penalizes model complexity
2. **VAF** (higher is better): Measures reconstruction accuracy
3. **Smoothness** (lower is better): Penalizes high-frequency noise in factor scores
4. **VSS** (higher is better): Rewards interpretable simple structure

Each metric is normalized to [0, 1], then combined:

```
score = w_BIC × BIC_norm + w_VAF × (1 - VAF_norm)
      + w_smooth × smooth_norm + w_VSS × (1 - VSS_norm)
```

The r with the minimum score (among those satisfying VAF threshold) is selected.

### Weight Configuration Guidelines

| Scenario | Recommended Weights | Rationale |
|----------|---------------------|-----------|
| Interpretation focus | `VSS = 2, others = 1` | Prioritize simple, interpretable loadings |
| Generalization focus | `BIC = 2, others = 1` | Avoid overfitting |
| EMG time-series | `smoothness = 2, others = 1` | Leverage temporal structure |
| Balanced (default) | All = 1 | Equal consideration |

```{r weight-examples}
# Example: Emphasize interpretability
result_interpret <- select_r_multicriteria_mfa(
  list_of_data, K = 3, rvec = 2:6,
  weights = c(BIC = 1, VAF = 1, smoothness = 1, VSS = 2)
)

# Example: Emphasize generalization
result_general <- select_r_multicriteria_mfa(
  list_of_data, K = 3, rvec = 2:6,
  weights = c(BIC = 2, VAF = 1, smoothness = 1, VSS = 1)
)

cat("Interpretation-focused r:", result_interpret$best_r, "\n")
cat("Generalization-focused r:", result_general$best_r, "\n")
```

## Understanding the Metrics

This section provides detailed explanations of each metric used in r-selection.

### Cross-Validation (CV)

**Purpose**: Estimate generalization performance by testing on held-out subjects.

**Method**: Leave-one-subject-out cross-validation

- For each subject i, fit model on remaining N-1 subjects
- Evaluate negative log-likelihood or Frobenius norm on held-out subject
- Average across all subjects

**When to use**: Primary criterion for global r estimation

**Interpretation**:
- Lower CV error = better generalization
- Look for "elbow" in the CV curve
- Selected r minimizes CV error

```{r cv-example}
# Direct CV-based selection
result_cv <- select_r_global_mfa(
  list_of_data,
  rvec = 2:6,
  criterion = "CV",
  cv_folds = 5,
  cv_metric = "NegLogLik"
)

cat("CV-selected r:", result_cv$r_global, "\n")
```

### Variance Accounted For (VAF)

**Purpose**: Measure how much variance the model explains.
$$
\text{VAF} = 1 - \frac{\sum_i \|X_i - \hat{X}_i\|^2}{\sum_i \|X_i - \bar{X}\|^2}
$$

**Interpretation**:
- VAF = 0.90 means 90% of variance explained
- Increases monotonically with r (more factors = more variance explained)
- Cannot detect overfitting alone

**Typical thresholds** (muscle synergy literature):
- Minimum acceptable: 0.80
- Good: 0.85-0.90
- Excellent: > 0.95

**Caution**: Use with other metrics; VAF alone favors larger r.

### Smoothness Penalty

**Purpose**: Detect overfitting by measuring temporal smoothness of factor scores.

**Method**: Sum of squared second-order differences
$$
\text{Smoothness} = \sum_{i,j,t} (z_{i,j,t} - 2z_{i,j,t-1} + z_{i,j,t-2})^2
$$

**Interpretation**:
- Low smoothness = smooth, physiologically plausible activations
- High smoothness = "wiggly" factors fitting high-frequency noise
- EMG-specific: leverages expected temporal structure

**When to use**: Particularly useful for EMG/muscle synergy analysis where factor scores should vary smoothly over time.

### Very Simple Structure (VSS)

**Purpose**: Evaluate interpretability of factor loadings.

**Concept**: Each muscle should load strongly on few (ideally one) synergy.

**Two methods**:

1. **Threshold method**: Proportion of muscles with exactly one loading > threshold
2. **Ratio method**: Average of max(λ²)/sum(λ²) per muscle

**Interpretation**:
- High VSS = each muscle clearly associated with one synergy
- Low VSS = complex, hard-to-interpret loadings

**Note**: Apply after varimax rotation for meaningful results:

```{r vss-example}
# Compute VSS with rotation
vss_rotated <- compute_vss_criterion(
  mfa_fit$Lambda,
  threshold = 0.5,
  method = "ratio",
  rotate = "varimax"
)
```

### Parallel Analysis (PA)

**Purpose**: Determine r by comparing eigenvalues to random data.

**Method**:
1. Compute eigenvalues of pooled covariance matrix
2. Generate random permutations of data
3. Compute 95th percentile of random eigenvalues
4. Select r = number of observed eigenvalues exceeding random threshold

**Interpretation**:
- Conservative estimate (often underestimates r)
- Good for exploratory analysis
- Less suitable when cluster structure is strong

```{r pa-example}
# Global MFA-based r estimation (replaces parallel analysis)
result_pa <- select_r_global_mfa(
  list_of_data,
  rvec = 1:8
)

cat("Selected r:", result_pa$r_global, "\n")
```

### Bayesian Information Criterion (BIC)

**Purpose**: Balance model fit and complexity.
$$
\text{BIC} = -2 \log L + p \log(n)
$$

where p = number of parameters, n = sample size.

**Interpretation**:
- Lower BIC = better model
- Automatically penalizes complexity
- Used in K,r grid search for joint selection

## Summary

The unified r-selection framework provides three complementary approaches:

| Mode | Speed | Stability | Flexibility | Recommended Use |
|------|-------|-----------|-------------|-----------------|
| **Global-first** | Fast | High | Low | Quick baseline, large datasets |
| **Clusterwise** | Moderate | Moderate | High | Heterogeneous populations |
| **Hybrid** | Slow | High | High | **Default recommendation** |

**Key takeaways:**

1. Use `select_r_mixture()` as the unified entry point
2. Start with `mode = "hybrid"` for most applications
3. Always visualize results with `plot(result)`
4. Consider cluster-specific estimates when interpreting results
5. Use `verbose = TRUE` to understand the selection process

## References

For more information on the underlying methods:

- **Global estimation**: See `?select_r_global_mfa` for CV and Parallel Analysis
- **Clusterwise evaluation**: See `?select_r_clusterwise_mfa` for metrics and aggregation
- **Mixture models**: See the main tutorial vignette for MFA/MPCA details

## Session Info

```{r session-info}
sessionInfo()
```
