---
title: 'Simulation Study: Effect of Cluster Separation on Model Performance'
author: "synergyMixR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study: Effect of Cluster Separation on Model Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "figures/sim-sep-",
  eval = FALSE
)

library(synergyMixR)
library(parallel)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# 1 Overview

This vignette replicates **Simulation 2** from the manuscript, in which we examine how
*between-cluster separation* ($\text{sep}$) influences the ability of

* **Mixture Factor Analysis (MFA)** and **Mixture PCA (MPCA)**
* **SingleFA** and **SinglePCA**

to recover latent sub-groups and reconstruct EMG signals.

The procedure varies the *separation hyper-parameter* $\text{sep}\in\{0.1,0.3,0.5,0.7,0.9,1.0\}$ while keeping
all other generative settings fixed (Table 1).  After fitting each candidate model, we record

* Bayesian Information Criterion (BIC),
* Adjusted Rand Index (ARI) against the true cluster labels, and
* Variance Accounted For (VAF).

# 2 Simulation design

| Parameter                  | Value                      | R argument        |
| -------------------------- | -------------------------- | ----------------- |
| Number of subjects         | $N=50$                     | `N_vec = 50`      |
| True clusters              | $K_{true}=4$               | `K_true_vec = 4`  |
| Rank per cluster           | $r_{true}=4$               | `r_true_vec = 4`  |
| Muscles (channels)         | $M=8$                      | `M_vec = 8`       |
| Time points                | $T=100$                    | `T_each = 100`    |
| Between-cluster separation | $\text{sep}=0.1\dots1.0$   | `sep_vec`         |
| Noise SD                   | $\sigma=2.0$               | `noise_vec = 2.0` |
| Candidate ranks            | $r\in[3,5]$                | `r_candidates`    |
| Candidate clusters         | $K\in[1,5]$                | `K_candidates`    |
| EM iterations              | 20                         | `max_iter`        |
| Random restarts            | 5                          | `n_rep_init`      |


> **Tip.**  Use fewer values or a smaller `max_iter` when prototyping to speed up the vignette build.


# 3 Running the simulation

```{r simulate}

set.seed(123)  # reproducibility of meta-parameters

result_dir <- "simulation_sep"

# This runs all combinations in parallel and saves .rds files
simulate_and_run_parallel_save(
  K_true_vec    = 4,

  r_true_vec    = 4,
  N_vec         = 50,
  M_vec         = 8,
  T_each        = 100,
  sep_vec       = seq(0.1, 1, 0.2),  # 0.1 to 1.0
  noise_vec     = 2.0,
  K_candidates  = 1:5,
  r_candidates  = 3:5,
  max_iter      = 20,
  n_rep_init    = 5,
  seed_data_base = 1000,
  mc.cores      = 10,

  output_dir    = result_dir
)

```



The function automatically uses `parallel::mclapply()` internally, storing one `.rds` file per scenario.

# 4 Aggregating results

After running the simulation, you can load and aggregate the results:

```{r load-results, message=FALSE, warning=FALSE}
result_dir <- "simulation_sep"  # adjust to your path

# Load all .rds files from the result directory
rds_files <- list.files(result_dir, pattern = "\\.rds$", full.names = TRUE)
df_raw <- do.call(rbind, lapply(rds_files, readRDS))

# Compute summary table
df_result <- compute_ari_vaf_table(df_raw) %>%
  mutate(across(c(sep, noise), as.factor))

head(df_result)
```

We reshape the data to *long* format for ggplot convenience:

```{r tidy}
df_long_ari <- df_result %>%
  select(K_true, r_true, N, M, sep, noise, ARI_MFA, ARI_PCA) %>%
  pivot_longer(c(ARI_MFA, ARI_PCA),
               names_to  = "Method",
               values_to = "ARI")
```

# 5 Visualisation

## 5.1 Clustering accuracy (ARI)

```{r ari-plot, fig.cap="ARI versus true cluster count, facetted by separation / noise."}
ggplot(df_long_ari,
       aes(x = factor(K_true), y = ARI, fill = Method)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  facet_wrap(~ paste("sep =", sep, " noise =", noise), nrow = 2) +
  theme_bw() +
  labs(title = "Adjusted Rand Index (ARI)",
       x = "True K", y = "ARI", fill = "Model")
```

## 5.2 Model fit (BIC & VAF)

```{r bic-vaf-plots, echo=FALSE}
# Example combined figure (optional)
```

> **Observation.** Higher separation ($\text{sep}\ge 0.5$) makes cluster boundaries more distinct, leading to ARI close to 1 for MFA and MPCA.  When clusters overlap ($\text{sep}=0.1$), both mixture models struggle, yet still outperform single-subject baselines.

# 6 Session information

```{r session, echo=FALSE, eval=TRUE}
sessionInfo()
```
